---
title: "2024 CABI PK/PD Data Analysis Exercise Part.3 - PK/PD relationship and PD model parameter estimation"
author: "Jin Gyu Kim"
date: ""
output: 
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='.', echo = TRUE, warning = FALSE)

rm(list=ls())
```

Welcome to PK data analysis exercise template, here we want to demonstrate how coding helps our analysis. 
Feel free to ask any questions or if you are having trouble running this script

# Install packages 
If Windows, click [here](https://cran.r-project.org/bin/windows/Rtools/) and download RTools with your R version. Check the console R version, it could be either 4.4.0, or 4.3.X

If mac, click [here](https://mac.r-project.org/tools/) and download GNU Fortran compiler to run this script on mac

# Load Packages
```{r echo=TRUE}
library(renv)
library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose4)
library(xpose.nlmixr2)
library(rxode2)
```

# PK/PD relationship
```{r warning = FALSE}
real_world_dataset <- read.csv("./dataset/analyzing_dataset.csv")

# Individual
ggplot(real_world_dataset[real_world_dataset$ID == 2, ]) +
  geom_line(data = real_world_dataset[real_world_dataset$ID == 2, ], aes(x = TIME, y = RESP), color = 'red') +
  geom_line(data = real_world_dataset[real_world_dataset$ID == 2, ], aes(x = TIME, y = DV), color = 'blue') +
  theme(legend.position = 'none') +
  scale_y_log10() +
  theme_bw()

# Population
ggplot(real_world_dataset[real_world_dataset$ID %in% c(1:12), ]) +
  geom_line(data = real_world_dataset[real_world_dataset$ID %in% c(1:12), ], aes(x = TIME, y = RESP), color = 'red') +
  geom_line(data = real_world_dataset[real_world_dataset$ID %in% c(1:12), ], aes(x = TIME, y = DV), color = 'blue') +
  facet_wrap(~ID) +
  theme(legend.position = 'none') +
  scale_y_log10() +
  theme_bw()

```

Legacy from PK estimation
```{r}
# Import PD observation dataset - Individual PK values
pk_run_fit <- read.csv('./dataset/pk_final_model_run.csv')
pk_run_fit

ncol(pk_run_fit)
nrow(pk_run_fit)

# 1 row per 1 ID and then have k20 k23 k32 ka vc
pk_run_fit_1 <- pk_run_fit %>% select(ID, ka, vc, k20, k23, k32)

pk_run_fit_2 <- pk_run_fit_1[!duplicated(pk_run_fit_1), ]

#xpdb_pk_cov <- xpose_data_nlmixr2(pk_run_fit,
#                                  xp_theme = theme_xp_nlmixr2())

#prm_distrib(xpdb_pk_cov)
#eta_distrib(xpdb_pk_cov)

pk_parameter_dataset <- data.frame(pk_run_fit) %>% select(ID, TIME, ka, vc, k20, k23, k32, MIC_C1D4, RESP)
indiv_pk_values <- pk_parameter_dataset[!duplicated(pk_parameter_dataset$ID), ]

ggplot(indiv_pk_values, aes(x = k20)) +
  geom_histogram(color = 'white') +
  theme_bw()

ggplot(indiv_pk_values, aes(x = k23)) +
  geom_histogram(color = 'white') +
  theme_bw()

ggplot(indiv_pk_values, aes(x = k32)) +
  geom_histogram(color = 'white') +
  theme_bw()

ggplot(indiv_pk_values, aes(x = vc)) +
  geom_histogram(color = 'white') +
  theme_bw()
```
# PD analysis
From PK analysis, we obtain individual PK parameters (e.g. k20, k23, k32, etc.) to predict the drug behavior (conc-time) when building a PD model 
```{r}
# Add dosing event row
event_row <- data.frame(ID = 1:100, TIME = 0, ka = NA, vc = NA, k20 = NA, k23 = NA, k32 = NA, MIC_C1D4 = NA, RESP = NA)

# Combine with RESP observation
pd_run_dataset <- rbind(pk_parameter_dataset, event_row)
pd_run_dataset <- pd_run_dataset %>% arrange(ID, TIME)
pd_run_dataset <- pd_run_dataset %>% group_by(ID) %>% fill(ka, vc, k20, k23, k32, MIC_C1D4, RESP, .direction = 'up')

# Clean dataset
pd_run_dataset <- pd_run_dataset %>% mutate(amt = ifelse(TIME == 0, 60000, 0),
                                            cmt = ifelse(TIME == 0, 1, 4),
                                            evid = ifelse(TIME == 0, 1, 0))
pd_run_dataset <- pd_run_dataset %>% select(ID, amt, TIME, evid, cmt, ka, vc, k20, k23, k32, MIC_C1D4, RESP) %>% ungroup()
pd_run_dataset <- pd_run_dataset %>% rename(DV = RESP)
```

## Parameter estimation
```{r warning = FALSE}
# We will build a pharmacokinetic model that describe the conc-time dataset
pd_model <- function() {
  ini({
    ## Theta - Initial estimates
    TVEMAX  <- log(10)
    TVEC50  <- log(200)
    TVKOUT  <- log(0.0314)
    TVHILL  <- log(2)
    ## Omega  - Initial estimates
    eta.EMAX ~ 0.1
    eta.EC50 ~ 0.1
    eta.KOUT ~ 0.1
    eta.HILL ~ 0.1
    ## Sigma - Initial estimates
    prop.err <- 0.05
    add.err  <- 0.2
  })
  model({
    Ka      <- ka
    EMAX    <- exp(TVEMAX + eta.EMAX)   # individual value of clearance
    EC50    <- exp(TVEC50 + eta.EC50)    # individual value of VC
    KOUT    <- exp(TVKOUT + eta.KOUT)
    HILL    <- exp(TVHILL + eta.HILL)
    
    ik20    <- k20
    ik23    <- k23
    ik32    <- k32
    
    MICBL   <- MIC_C1D4
    Kin     <- KOUT * MICBL
    
    RESP(0) =  MICBL
    
    d/dt(absorption) = -Ka*absorption
    d/dt(central)    = Ka*absorption - ik20 * central - ik23*central + ik32*peripheral
    d/dt(peripheral) = ik23*central - ik32*peripheral 
    
    CP = central / vc       # concentration in plasma
    
    EFFECT = 1 + EMAX * CP^HILL / (EC50^HILL + CP^HILL)
    
    d/dt(RESP)       = Kin*EFFECT - KOUT*RESP
  
    RESP ~ prop(prop.err) + add(add.err)    # define error model
  })
}

# We incorporate the model with the dataset to obtain pharmacokinetic parameters
pd_fit <- nlmixr2(pd_model, 
                  pd_run_dataset, 
                  "focei",
                  control=list(print=1), 
                  table=list(cwres=TRUE, npde=TRUE)
                  )
```   


## Model Diagnosis
We visualize the statistical output since seeing the plot is better than just reading the numbers 
```{r warning = FALSE, echo = FALSE}
# We will convert 
xpdb_pd <- xpose_data_nlmixr2(pd_fit,
                           xp_theme = theme_xp_nlmixr2())

# Individual plots 
ind_plots(xpdb_pd)

# observation vs population prediction plot
dv_vs_pred(xpdb_pd, type="ps")

# observation vs individual prediction plot
dv_vs_ipred(xpdb_pd, type="ps")

# Residual error vs time plot
res_vs_idv(xpdb_pd, res="CWRES", type="ps")

# individual prediction vs individual
ipred_vs_idv(xpdb_pd, group="ID")

# Parameter distribution
#eta_distrib(xpdb_pd)

# Residual distribution
#res_distrib(xpdb_pd, res="CWRES")
```

```{r}
individual_parameters <- pd_fit %>% select(ID, EMAX, EC50, KOUT, HILL, MICBL, k20, k23, k32, vc)
individual_parameters <- individual_parameters %>% distinct(ID, .keep_all = TRUE)

write.csv(individual_parameters, './dataset/individual_pkpd_param.csv', row.names = FALSE)
```

