---
title: "2024 CABI PK/PD Data Analysis Exercise Part.4 - Simulation Run and Precision dosing"
author: "Jin Gyu Kim"
date: ""
output: 
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---
```{r setup, include=FALSE}
rm(list=ls())
```

Welcome to PK/PD data analysis exercise template, here we want to demonstrate how coding helps our analysis. 
Feel free to ask any questions or if you are having trouble running this script

# Install packages 
If Windows, click [here](https://cran.r-project.org/bin/windows/Rtools/) and download RTools with your R version. Check the console R version, it could be either 4.4.0, or 4.3.X

If mac, click [here](https://mac.r-project.org/tools/) and download GNU Fortran compiler to run this script on mac

# Load Packages
Here we want to import our installed package into our R environment
```{r echo=TRUE}
#install.packages("renv")
library(renv)
#renv::init()

library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose.nlmixr2)
library(rxode2)
library(mrgsolve)
```

Read and clean data for simulation 
```{r pressure, echo=TRUE}
# Import data
patient_dataset <- read.csv("./dataset/individual_pkpd_param.csv")
mod <- mread("model/simulation_model.mod")
```

```{r}
e <- ev(ID = 1:100, amt = 60000, time = seq(0, 24*30, by = 24))
individual_sim_dataset <- merge(patient_dataset, e, by = 'ID')

set.seed(100)
sim <- mod %>% ev(individual_sim_dataset) %>% mrgsim_df(tgrid = tgrid(0, 720, delta = 1), carry_out = c('k20', 'k23', 'k32', 'MICBL', 'vc'))
```

```{r}
# PK simulation
ggplot(sim[sim$ID %in% c(1:12) & sim$time < 24,  ], aes(x = time/24, y = DV_PK, group = ID, color = as.character(ID))) +
  geom_line() +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(x = 'Time (Days)', y = 'Conc (mg/L)')

ggplot(sim[sim$ID %in% c(1:12), ], aes(x = time/24, y = DV_PK, group = ID, color = as.character(ID))) +
  geom_line() +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(x = 'Time (Days)', y = 'Conc (mg/L)')

ggplot(sim[sim$ID %in% c(1:12) & sim$time < 24,  ], aes(x = time/24, y = DV_RESP, group = ID, color = as.character(ID))) +
  geom_line() +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(x = 'Time (Days)', y = 'Conc (mg/L)')

ggplot(sim[sim$ID %in% c(1:12),  ], aes(x = time/24, y = DV_RESP, group = ID, color = as.character(ID))) +
  geom_line() +
  theme_bw() +
  theme(legend.position = 'none')+
  labs(x = 'Time (Days)', y = 'Conc (mg/L)')
  
ggplot() +
  geom_line(data = sim[sim$ID %in% c(1:12) & sim$time < 24, ], aes(x = time, y = DV_PK, group = ID), color = 'red') +
  geom_line(data = sim[sim$ID %in% c(1:12) & sim$time < 24, ], aes(x = time, y = DV_RESP, group = ID), color = 'blue') +
  theme_bw() +
  facet_wrap(~ID) +
  labs(x = 'Time (Days)', y = 'Conc (mg/L)')

ggplot() +
  geom_line(data = sim[sim$ID %in% c(1:12), ], aes(x = time, y = DV_RESP, group = ID), color = 'blue') +
  geom_line(data = sim[sim$ID %in% c(1:12), ], aes(x = time, y = DV_PK, group = ID), color = 'red') +
  theme_bw() +
  facet_wrap(~ID) +
  labs(x = 'Time (Days)', y = 'Conc (mg/L)')
```

# Set the therapeutic range from 100 - 200 mg/L
```{r}
ggplot(data = sim[sim$ID %in% c(50:62), ], aes(x = time/24, y = DV_RESP, group = ID, color = as.character(ID))) +
  geom_line(alpha = 0.5) +
  annotate('text', x = 27, y = 90, label = 'Below therapeutic range', color = 'blue') +
  geom_hline(yintercept = 100, alpha = 1, color = 'blue') +
  geom_hline(yintercept = 200, alpha = 1, color = 'red') +
  annotate('text', x = 27, y = 210, label = 'Toxicity threshold', color = 'red') +
  annotate('rect', xmin = 0, xmax = 30, ymin = 100, ymax = 200, label = 'Therapeutic range', fill = 'yellow', alpha = 0.5) +
  annotate('text', x = 27, y = 150, label = 'Therapeutic range', color = 'purple') +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(title = '30 day simulation with 10 individuals', x = 'Time (Days)', y = 'Conc (mg/L)')

ggplot(data = sim[sim$ID %in% c(50:62), ], aes(x = time/24, y = DV_RESP, group = ID, color = as.character(ID))) +
  geom_line(alpha = 1) +
  geom_hline(yintercept = 100, alpha = 1, color = 'blue') +
  geom_hline(yintercept = 200, alpha = 1, color = 'red') +
  annotate('rect', xmin = 0, xmax = 30, ymin = 100, ymax = 200, label = 'Therapeutic range', fill = 'yellow', alpha = 0.5) +
  theme_bw() +
  theme(legend.position = 'none') +
  facet_wrap(~ID)

```

# Adjust the dosing for individuals
```{r}
# 50, 57, 58, 61 - Reduce dose
reduce_dose <- patient_dataset %>% filter(ID %in% c(50, 57, 58, 61))
e1 <- ev(ID = c(50, 57, 58, 61), amt = 30000, time = seq(0, 24*30, by = 24))
reduced_patients <- merge(reduce_dose, e1, by = 'ID')

sim_50_57_58_61 <- mod %>% ev(reduced_patients) %>% mrgsim_df(tgrid = tgrid(0, 720, delta = 1), carry_out = c('k20', 'k23', 'k32', 'MICBL', 'vc'))

ggplot(sim_50_57_58_61, aes(x = time/24, y = RESP, group = factor(ID), color = factor(ID))) +
  geom_line(alpha = 0.5) +
  #annotate('text', x = 27, y = 90, label = 'Below therapeutic range', color = 'blue') +
  geom_hline(yintercept = 100, alpha = 1, color = 'blue') +
  #annotate('text', x = 27, y = 150, label = 'Therapeutic range') +
  geom_hline(yintercept = 200, alpha = 1, color = 'red') +
  #annotate('text', x = 27, y = 210, label = 'Toxicity threshold', color = 'red') +
  annotate('rect', xmin = 0, xmax = 30, ymin = 100, ymax = 200, label = 'Therapeutic range', fill = 'yellow', alpha = 0.5) +
  theme_bw() +
  theme(legend.position = 'none') +
  facet_wrap(~ID) +
  labs(title = '30 day simulation with 50, 57, 58, 61', x = 'Time (Days)', y = 'Conc (mg/L)')
  
# 52, 55, 59 - Increase
increase_dose <- patient_dataset %>% filter(ID %in% c(52, 55, 59))
e2 <- ev(ID = c(52, 55, 59), amt = 120000, time = seq(0, 6*120, by = 24))
increased_patients <- merge(increase_dose, e2, by = 'ID')

sim_52_55_59 <- mod %>% ev(increased_patients) %>% mrgsim_df(tgrid = tgrid(0, 720, delta = 1), carry_out = c('k20', 'k23', 'k32', 'MICBL', 'vc'))

ggplot(sim_52_55_59, aes(x = time/24, y = RESP, group = factor(ID), color = factor(ID))) +
  geom_line(alpha = 0.5) +
  #annotate('text', x = 27, y = 90, label = 'Below therapeutic range', color = 'blue') +
  geom_hline(yintercept = 100, alpha = 1, color = 'blue') +
  #annotate('text', x = 27, y = 150, label = 'Therapeutic range') +
  geom_hline(yintercept = 200, alpha = 1, color = 'red') +
  #annotate('text', x = 27, y = 210, label = 'Toxicity threshold', color = 'red') +
  annotate('rect', xmin = 0, xmax = 30, ymin = 100, ymax = 200, label = 'Therapeutic range', fill = 'yellow', alpha = 0.5) +
  theme_bw() +
  theme(legend.position = 'none') +
  facet_wrap(~ID) +
  labs(title = '30 day simulation with 52, 55, 59', x = 'Time (Days)', y = 'Conc (mg/L)')
```

# Using Interactive Tool
Click [here](bimvptest.shinyapps.io/precision_dosing/) for running interactive tool = 

