---
title: "2024 CABI PK/PD Data Analysis Exercise Part.2 - Identifying patient characteristics driving drug disposition (PK) - Covariate building step"
author: "Jin Gyu Kim"
date: ""
output: 
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment='.', echo = TRUE, warning = FALSE)
rm(list=ls())
```

Welcome to PK/PD data analysis exercise template, here we want to demonstrate how coding helps our analysis. 
Feel free to ask any questions or if you are having trouble running this script

# Install packages 
If Windows, click [here](https://cran.r-project.org/bin/windows/Rtools/) and download RTools with your R version. Check the console R version, it could be either 4.4.0, or 4.3.X

If mac, click [here](https://mac.r-project.org/tools/) and download GNU Fortran compiler to run this script on mac

# Load Packages
Here we will import our installed packages into our R environment
```{r echo=TRUE}
library(renv)
library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose4)
library(xpose.nlmixr2)
library(rxode2)
```

# Build base PK model
```{r}
real_world_dataset <- read.csv("./dataset/analyzing_dataset.csv")

base_model <- function() {
  ini({
    ## Theta - Initial estimates
    tvka  <- log(0.479)
    tvcl  <- log(12.1)
    tvvc  <- log(67.9)
    tvvp  <- log(515)  
    tvq   <- log(25)
    #covWTVC  <- 0.75
    #covAgeCL <- 0.75
    ## Omega  - Initial estimates 
    eta.cl ~ 0.1
    eta.vc ~ 0.1
    eta.vp ~ 0.1
    eta.q  ~ 0.1
    ## Sigma - Initial estimates
    log.err <- 0.111 
    #add.err  <- 0.024   
  })
  model({
    ka <- exp(tvka)
    cl <- exp(tvcl + eta.cl)
    vc <- exp(tvvc + eta.vc)
    vp <- exp(tvvp + eta.vp)
    q  <- exp(tvq + eta.q)
    sex <- GENDER 
    baseline <- MIC_C1D4
    pk_obs <- RESP
    
    k20 <- cl/vc
    k23 <- q/vc
    k32 <- q/vp
    
    d/dt(absorption) = - ka * absorption
    d/dt(central)    = ka * absorption - k20 * central - k23 * central + k32 * peripheral        # model differential equation
    d/dt(peripheral) = k23 * central - k32 * peripheral 
    
    DV = log(central / vc)       # concentration in plasma
    DV ~ add(log.err)            # define error model
  })
}

# We incorporate the model with the dataset to obtain pharmacokinetic parameters
base_pk_fit <- nlmixr2(base_model, 
               real_world_dataset, 
               "focei",
              control=list(print=1), 
              table=list(cwres=TRUE, npde=TRUE)
              )
```

```{r}
# Output
base_pk_fit

# We will convert 
xpdb_pk <- xpose_data_nlmixr2(base_pk_fit,
                           xp_theme = theme_xp_nlmixr2())

# individual prediction vs individual
pred_vs_idv(xpdb_pk, group="ID")

# individual plots 
ind_plots(xpdb_pk)

# observation vs individual prediction plot
dv_vs_pred(xpdb_pk, type="ps")

# residual error vs population prediction plot
res_vs_pred(xpdb_pk, res="CWRES", type="ps")

# residual error vs individual plot
res_vs_idv(xpdb_pk, res="CWRES", type="ps")
```



# Load dataset and initial screen
```{r, echo = TRUE}
# Import our analyzing dataset with demographic data
dataset <- real_world_dataset %>% mutate(GENDER = ifelse(GENDER == 1, 'Male', 'Female'))

# Spaghetti plot for all patients
ggplot(dataset, aes(x = TIME, y = DV, color = factor(ID))) +
  geom_line() + 
  theme_bw() +
  #scale_y_log10() +
  theme(legend.position = 'none') +
  labs(title = "Spaghetti Plot: Conc vs Time curve from 100 individuals", x = "Time (hr)", y = "Conc (mg/L)")

# Spaghetti plot for all patients
ggplot(dataset[dataset$ID %in% c(1:12), ], aes(x = TIME, y = DV, color = factor(ID))) +
  geom_line() + 
  geom_point() +
  theme_bw() +
  #scale_y_log10() +
  facet_wrap(~ID) +
  theme(legend.position = 'none') +
  labs(title = "Spaghetti Plot: Conc vs Time curve from 100 individuals", x = "Time (hr)", y = "Conc (mg/L)")

# Since we have multiple observations from every patients, we will filter only one row per patient
dataset_ID_unique <- dataset[!duplicated(dataset$ID), ]
base_pk_param <- base_pk_fit[!duplicated(base_pk_fit$ID), ]

pk_covariate_exploration <- merge(dataset_ID_unique, base_pk_param, by = 'ID')

# Summary of patient demographics
summary(dataset_ID_unique)
```

# BSV Exploration: Gender 
```{r, echo = TRUE, warning = FALSE}
ggplot(dataset_ID_unique, aes(x = GENDER, fill = GENDER)) +
  geom_bar() +
  theme_bw() +
  labs(title = "Distribution of Gender")

# One plot
ggplot(dataset, aes(x = TIME, y = DV, group = ID, color = GENDER)) +
  geom_line() + 
  #geom_point() +
  theme_bw() +
  scale_y_log10() +
  labs(title = "Spaghetti Plot: Stratified by Gender", x = "Time (hr)", y = "Conc (mg/L)")
  #theme(legend.position = 'none') 

# Facet
ggplot(dataset, aes(x = TIME, y = DV, group = ID, color = GENDER)) +
  geom_line() + 
  #geom_point() +
  theme_bw() +
  scale_y_log10() +
  facet_wrap(~GENDER) +
  theme(legend.position = 'none') +
  labs(title = "Spaghetti Plot: Stratified by Gender", x = "Time (hr)", y = "Conc (mg/L)")

# Exploring gender factor in individual PK component. 
ggplot(pk_covariate_exploration, aes(x = GENDER.x, y = eta.cl)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Gender', y = 'Individual Clearance (CL)')

ggplot(pk_covariate_exploration, aes(x = GENDER.x, y = eta.vc)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Gender', y = 'Individual Central Volume (VC)')

ggplot(pk_covariate_exploration, aes(x = GENDER.x, y = eta.vp)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Gender', y = 'Individual Peripheral Volume (VP)')

ggplot(pk_covariate_exploration, aes(x = GENDER.x, y = eta.q)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Gender', y = 'Individual Intercompartmental Clearance (Q)')
```

## BSV Exploration: Age
```{r, echo = TRUE, warning = FALSE}

ggplot(dataset_ID_unique, aes(x = Age)) +
  geom_histogram(color = 'white',
                 bins = 10,
                 center = 4) +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(title = "Distribution of Age", x = 'Age (yr)')

# One plot
ggplot(dataset, aes(x = TIME, y = DV, group = ID, color = Age > 30)) +
  geom_line() + 
  #geom_point() +
  theme_bw() +
  scale_y_log10() +
  labs(title = "Spaghetti Plot: Stratified by Age", x = "Time (hr)", y = "Conc (mg/L)")
  #theme(legend.position = 'none') 

# Facet
ggplot(dataset, aes(x = TIME, y = DV, group = ID, color =  Age > 30)) +
  geom_line() + 
  #geom_point() +
  theme_bw() +
  scale_y_log10() +
  facet_wrap(~ Age > 30) +
  labs(title = "Spaghetti Plot: Stratified by Age", x = "Time (hr)", y = "Conc (mg/L)")
  #theme(legend.position = 'none') 

# Exploring age factor in individual PK component. 
# Evident that Age factor relates to individual Cl. Strengthening hypothesis from the prelim study 
ggplot(pk_covariate_exploration, aes(x = Age, y = eta.cl)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Age', y = 'Individual Clearance (CL)')

ggplot(pk_covariate_exploration, aes(x = Age, y = eta.vc)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Age', y = 'Individual Central Volume (VC)')

ggplot(pk_covariate_exploration, aes(x = Age, y = eta.vp)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Age', y = 'Individual Peripheral Volume (VP)')

ggplot(pk_covariate_exploration, aes(x = Age, y = eta.q)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Age', y = 'Individual Intercompartmental Clearance (Q)')
```

## BSV Exploration: Weight
```{r, echo = TRUE, warning = FALSE}

# Take a look at Age.yr, Weight, and BMI but this isn't enough to understand the population characteristics
ggplot(dataset_ID_unique, aes(x = WT)) +
  geom_histogram(color = 'white',
                 bins = 20,
                 center = 4) +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(title = "Distribution of Weight", x = "Weight (kg)")

# One plot
ggplot(dataset, aes(x = TIME, y = DV, group = ID, color = WT < 60)) +
  geom_line() + 
  #geom_point() +
  theme_bw() +
  scale_y_log10() +
  labs(title = "Spaghetti Plot: Stratified by Weight", x = "Time (hr)", y = "Conc (mg/L)")

# Facet
ggplot(dataset, aes(x = TIME, y = DV, group = ID, color = WT < 60)) +
  geom_line() + 
  #geom_point() +
  theme_bw() +
  scale_y_log10() +
  facet_wrap(~WT < 60) +
  labs(title = "Spaghetti Plot: Stratified by Weight", x = "Time (hr)", y = "Conc (mg/L)")

# Exploring weight factor in individual PK component.
ggplot(pk_covariate_exploration, aes(x = WT, y = eta.cl)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Weight', y = 'Individual Clearance (CL)')

# Evident that WT factor relates to individual VC. Strengthening hypothesis from the prelim study 
ggplot(pk_covariate_exploration, aes(x = WT, y = eta.vc)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Weight', y = 'Individual Central Volume (VC)')

ggplot(pk_covariate_exploration, aes(x = WT, y = eta.vp)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Weight', y = 'Individual Peripheral Volume (VP)')

ggplot(pk_covariate_exploration, aes(x = WT, y = eta.q)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = 'Weight', y = 'Individual Intercompartmental Clearance (Q)')

```

#PK model incorporating patient characteristics
```{r}
# Age in Clearance and WT in Volume
covariate_model <- function() {
  ini({
    ## Theta - Initial estimates
    tvka  <- log(0.479)
    tvcl  <- log(12.1)
    tvvc  <- log(67.9)
    tvvp  <- log(515)  
    tvq   <- log(25)
    covWTVC  <- 0.75
    covAgeCL <- 0.75
    ## Omega  - Initial estimates 
    #eta.ka ~ 0.1
    eta.cl ~ 0.1
    eta.vc ~ 0.1
    eta.vp ~ 0.1
    eta.q  ~ 0.1
    ## Sigma - Initial estimates
    log.err <- 0.111 
    #add.err  <- 0.024   
  })
  model({
    ka <- exp(tvka)
    cl <- exp(tvcl + eta.cl) * (Age/30)^covAgeCL   # individual value of clearance
    vc <- exp(tvvc + eta.vc) * (WT/70)^covWTVC    # individual value of VC
    vp <- exp(tvvp + eta.vp)
    q  <- exp(tvq + eta.q)
    sex <- GENDER 
    baseline <- MIC_C1D4
    pk_obs <- RESP
    
    k20 <- cl/vc
    k23 <- q/vc
    k32 <- q/vp
    
    d/dt(absorption) = - ka * absorption
    d/dt(central)    = ka * absorption - k20 * central - k23 * central + k32 * peripheral     # model differential equation
    d/dt(peripheral) = k23 * central - k32 * peripheral 
    
    DV = log(central / vc)       # concentration in plasma
    DV ~ add(log.err)            # define error model
  })
}

pk_covariate_fit <- nlmixr2(covariate_model, 
               real_world_dataset, 
               "focei",
              control=list(print=1), 
              table=list(cwres=TRUE, npde=TRUE))
```

# Compare fits between base model and covariate model
```{r}
print(base_pk_fit)
print(pk_covariate_fit)

rbind(base_pk_fit$objDf, pk_covariate_fit$objDf)

# We will convert 
xpdb_pk_cov <- xpose_data_nlmixr2(pk_covariate_fit,
                           xp_theme = theme_xp_nlmixr2())

# individual prediction vs individual
pred_vs_idv(xpdb_pk) # 1st
pred_vs_idv(xpdb_pk_cov, group="ID") # 2nd

# Individual plots 
ind_plots(xpdb_pk_cov)

# observation vs individual prediction plot
dv_vs_pred(xpdb_pk, type="ps")
dv_vs_pred(xpdb_pk_cov, type="ps")

# Residual error vs population prediction plot
res_vs_pred(xpdb_pk_cov, res="CWRES", type="ps")

# Residual error vs individual plot
res_vs_idv(xpdb_pk_cov, res="CWRES", type="ps")

write.csv(pk_covariate_fit, './dataset/pk_final_model_run.csv', row.names = FALSE)

# Residual distribution
#res_distrib(xpdb_pk_cov, res="CWRES")
```

### Bonus: BOV Exploration
```{r, echo = TRUE, warning = FALSE}
iov_dataset <- read.csv("./dataset/multiple_occasion.csv")

ggplot(iov_dataset[iov_dataset$ID == 1 & iov_dataset$time < 24, ], aes(x = time, y = CP, color = factor(ID))) +
  geom_line() +
  geom_point() +
  #scale_y_log10() +
  theme_bw() +
  #facet_wrap(~ID) +
  theme(legend.position = 'none') +
  labs(title = 'Individual conc-TIME profile with single dose', x = 'Time (Days)', y = 'Conc (mg/L)')

ggplot(iov_dataset[iov_dataset$ID == 1, ], aes(x = time, y = CP, color = factor(ID))) +
  geom_line() +
  geom_point() +
  theme_bw() +
  theme(legend.position = 'none') +
  labs(title = 'Individual conc-TIME profile with multiple dose: Intra-Occasion Variability', x = 'Time (hours)', y = 'Conc (mg/L)')
```
