---
title: "2025 CaBi DS Camp - Understanding basic PK/PD concepts and applying in personalized therapy: TEST"
author: "Jin Gyu Kim"
date: "2025-05-30"
output:
  html_document:
    css: docs/src/styles/styles.css
    toc: true
    number_sections: true
---

***
1. **Install the C Compiler**  
Ensure that a C compiler is installed on your system. For installation instructions, refer to `ReadMe.txt`.

2. **Activate the R environment**  
Run the following code chunk to install and activate the R project. When prompted in the console, type Y at install.packages()
```{r}
install.packages("renv")
library(renv)
renv::activate()
```

2-1. **Restore packages**
This step will allow you to sync the packages from renv.lock file
```{r, echo = TRUE, message = "hide", warning = "hide", results = "hide"}
renv::restore()
```

3. **Run Script**
To knit the `.Rmd` document, press `Ctrl + Shift + K` (Windows) or `Cmd + Shift + K` (Mac) in RStudio. This will generate the output document based on the settings defined in the YAML header. 

4. **Validate the Run**
`test.html` is an outcome after successful run from `test.Rmd` script. Compare with `test_run_complete.pdf` if anything is missing.
Compare this file with test_run_complete.pdf to ensure that no content is missing. If any issues are encountered, please contact the author for troubleshooting.

***

**Load Packages**
```{r echo = TRUE, message = "hide", warning = "hide", results = "hide"}
rm(list=ls())
library(tidyverse)
library(ggplot2)
library(nlmixr2)
library(xpose4)
library(xpose.nlmixr2)
library(gridExtra)
library(ggPMX)
library(ggpubr)
library(mrgsolve)
```

**Data Import**
```{r, warning = "hide"}
# Import busulfan two compartment dataset
test_dataset <- read.csv("dataset/analyzing_dataset.csv", na.strings = ".")

# Plasma drug concentration decline in two phases shown in semi-log plot
ggplot(busulfan_dataset, aes(TIME/60, log(DV), group = ID, color = as.factor(ID))) +
  geom_line() +
  theme_bw() +
  labs(title = "Test semi-log plot", x = "Time (hours)", y = "Conc (mg/L)") +
  theme(legend.position = "blank")
```

**Fit**
```{r, results = "hide"}
# One compartment model structure
test_model <- function() {
  ini({
    ## Theta - Initial estimates
    tvka  <- log(0.479)
    tvcl  <- log(12.1)
    tvvc  <- log(67.9)
    tvvp  <- log(515)  
    tvq   <- log(25)
    #covWTVC  <- 0.75
    #covAgeCL <- 0.75
    ## Omega  - Initial estimates 
    eta.cl ~ 0.1
    eta.vc ~ 0.1
    eta.vp ~ 0.1
    eta.q  ~ 0.1
    ## Sigma - Initial estimates
    log.err <- 0.111 
    #add.err  <- 0.024   
  })
  model({
    ka <- exp(tvka)
    cl <- exp(tvcl + eta.cl)
    vc <- exp(tvvc + eta.vc)
    vp <- exp(tvvp + eta.vp)
    q  <- exp(tvq + eta.q)
    sex <- GENDER 
    baseline <- MIC_C1D4
    pk_obs <- RESP
    
    k20 <- cl/vc
    k23 <- q/vc
    k32 <- q/vp
    
    d/dt(absorption) = - ka * absorption
    d/dt(central)    = ka * absorption - k20 * central - k23 * central + k32 * peripheral        # model differential equation
    d/dt(peripheral) = k23 * central - k32 * peripheral 
    
    DV = log(central / vc)       # concentration in plasma
    DV ~ add(log.err)            # define error model
  })
}

# 1 compartment model fitting
test_fit <- nlmixr2(
  test_model,
  test_dataset,
  "focei",
  table = list(cwres = TRUE)
)
```

**Plot**
```{r, warning = "hide"}
# Convert to XPOSE for goodness of fit
test_xpdb_pk <- xpose_data_nlmixr2(test_fit,
                                     xp_theme = theme_xp_nlmixr2()
                                     )

# observation vs population prediction plot
dv_vs_pred(test_xpdb_pk,
           type="pls",
           title = "Test",
           log   = 'xy',
           subtitle = NULL,
           guide = TRUE)
```
